A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-30 paper.tex
Create "Full ARG" section

% TODO update this when later sections have been drafted.
As we see in sections XXX and YYY there are substantial limitations
of expressivity and computational performance when using this eARG
encoding. More fundamentally, however, the general approach
of classifying nodes according to a discrete set of events
and associating genome information with those nodes is intrinsically
limiting. As we saw in the previous section, the original ARG
definitions were derived directly from the coalescent with recombination
stochastic process, and the types of ancestry that can be
described are therefore limited by the assumptions of that
model. While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.

% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$. This
assumption ensures that multiple events (i.e., simultaneous
coalescence and recombination) don't occur at the same
time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.

Another assumption of the coalescent with recombination which prevents complex
combinations of events occurring simultaneously, is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
Clearly multiple recombinations (and other complex events) will
occur in the ancestry of such datasets, and we should not be
limited in what we can infer by the data structure that we
use to represent the results.

@@

@@

@@

\item Comparison metrics are a problem. All evaluations has essentially
looked at either some form of pairwise comparison of trees along the genome
using tree metrics, or looked at the distributions of TMRCAs. No
one has succeeded in performing any comparisons of ARGs beyond
regarding them as a sequence of independent trees. If we genuinely believe
that ARGs are more than this, then we need to devise some metrics that
consider more than the local tree structure.

@@

TODO mention \citep{schaefer2021ancestral} somewhere.

>>>>

% TODO update this when later sections have been drafted.
% As we see in sections XXX and YYY there are substantial limitations
% of expressivity and computational performance when using this eARG
% encoding. More fundamentally, however, the general approach
% of classifying nodes according to a discrete set of events
% and associating genome information with those nodes is intrinsically
% limiting. As we saw in the previous section, the original ARG
% definitions were derived directly from the coalescent with recombination
% stochastic process, and the types of ancestry that can be
% described are therefore limited by the assumptions of that
% model.

@@

\section*{When do we need ``full'' ARGs?}
% The full ancestral recombination graph (ARG) is a structure that encodes all
% coalescence and recombination events resulting from the stochastic process of
% the coalescent with recombination.
\citet{brandt2021evaluation} define a ``full'' ARG as
``a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination'' (thus going against the general
historial trend discussed in the XXX section).
\citet{rasmussen2022espalier} have a less restrictive definition, in
which a ``full'' ARG is one that contains complete information about
recombination events, without necessarily being tied to a particular
stochastic process.
While [something positive], it is important to be clear on
what the overall goal of such precise inferences are, and
on the realistic limitations are on when they can be made.

The main reasons for wanting precise estimates of all recombination
events are to either compute a likelihood of the inferred ARG
under a statistic model, or to study the recombinants in
detail~\citet{rasmussen2022espalier}.
% Explain that we're really interested in the details of recombination
% itself with a small number of samples.
In the latter case, we might
have a small number of samples~\citep{guo2022recombination}.

% If we want to use the coalescent, then really the assumptions
% need to make sense.
While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$. This
assumption ensures that multiple events (i.e., simultaneous
coalescence and recombination) don't occur at the same
time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.

Another assumption of the coalescent with recombination which prevents complex
combinations of events occurring simultaneously, is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
% Clearly multiple recombinations (and other complex events) will
% occur in the ancestry of such datasets, and we should not be
% limited in what we can infer by the data structure that we
% use to represent the results.

Such datasets cannot be reasonably be described by
the coalescent as a statical model, and so more non-parametric
methods like \tsinfer\ and \relate\ would seem more appropriate.
[More]

@@

\item In general, discussions tend to contrast an ARG
on one hand as holding \emph{complete} information about recombination
(and therefore between-tree correlation structure)
and a sequence of local trees on the other (with, by implication,
no direct information about between-tree correlations).
A key goal of this article is to show that this is a false dichotomy:
a \emph{continuum} of information about recombination and between-tree
correlation structure can be represented and inferred, and all
such levels of information are useful, even if only for computational
efficiency.

@@

\item Comparison metrics are a problem. All evaluations has essentially
looked at either some form of pairwise comparison of trees along the genome
using tree metrics as proposed by \citet{kuhner2015assessing},
or looked at the distributions of TMRCAs. No
one has succeeded in performing any comparisons of ARGs beyond
regarding them as a sequence of independent trees. If we genuinely believe
that ARGs are more than this, then we need to devise some metrics that
consider more than the local tree structure.

@@

TODO mention \citep{schaefer2021ancestral} somewhere.
Also \citep{kuhner2017consensus}.
